---
title: "My data summary"
author: "Mari J"
date: "Aug. 6, 2025"

categories: [goals,general]
---

**Introduction**: Based on the results of a 2022 study published by the [NORC at The University of Chicago](https://www.norc.org/research/library/survey--americans-see-cost-as-the-biggest-barrier-to-higher-educ.html), more than half of adults within the United States reported their beliefs that people are not attending college due to its excessive costs. These results seem to raise some concerns and questions regarding what type (if any) impact college tuition costs have had on unemployment rates

**Methods**:

I utilized 2 different data sets, one from the "[US Bureau of Labor Statistics](https://www.bls.gov/web/laus/laumstrk.htm)", listing a data set of unemployment rates by state. Additionally, I found a data set from "[Kaggle](https://www.kaggle.com/datasets/kfoster150/avg-cost-of-undergrad-college-by-state)" on the costs of undergrad tuition by state and type of institution. I utilized my filtering and grouping functions only to include the data of "4-year" institutions in my results

Here is my code.

```{r}
library(ggplot2)
library(tidyverse)

unemploymentrates_by_state <- read.csv("unemploymentrates.by.state.csv")
nces330_20 <- read.csv("nces330_20.csv")

nces330_20$unemployment <- unemploymentrates_by_state[match(nces330_20$State, unemploymentrates_by_state$State), 2]

nces330_20 <- nces330_20 |>
  filter(
    Expense == "Fees/Tuition" & Length == "4-year"
) 

ggplot(
  data = nces330_20,
  mapping = aes(y = State, x = Value)
) + stat_summary(
  fun.min = min,
  fun.max = max,
  fun = mean
)

```

**Discussion:**

These specific functions grouped both of my data sets together, by adding on columns of the unemployment rates into my "ncess330_20.csv" (undergrad tuition data) data set.

The graph shown above displays the "Values", costs of tuition on the x-axis, with the states being represented on the y-axis. The results displayed on this graph are represented by scale with the dot being farthest to the left if the unemployment rate is low and farthest to the right with higher unemployment rates.

**Results:**

Based on the data displayed in this graph, I can infer that most states typically show midpoint results for unemployment rates. Although, I noticed a very small pattern the states with tuition values that do not exceed 30,000 typically have a slightly lower unemployment rate, while a few states with higher tuition rates have a dot placement on the right side of the median.

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

### Data Separated by Institution Type

```{r}

nces330_20$unemployment_type <- "Low"
nces330_20[nces330_20$unemployment > 3.85, ]$unemployment_type <- "High"

ggplot(
  data = nces330_20,
  mapping = aes(x = unemployment, y = Value, color = unemployment_type)
) + geom_point(alpha = 0.5, position = "jitter") + facet_wrap(~Type) + geom_smooth(method = "lm") + scale_color_manual(values = c("Low" = "violet", "High" = "hotpink"))
```

**Method:**

For this graph listed above, I created variables called "Low" and "High" to separates the vales of the unemployment with the lowest being ***1.8*** and the highest being ***5.9***. I separated these by the median number of ***3.85***. I created a scatter plot chart using the "ggplot" function and then adding lines, using the "geom_smooth" function. Lastly, I changed the colors of the dots to violet and pink.

**Discussion:**

In this graph, I separated the data from my "Type" column, to see the results of unemployment rates based off the "values" (tuition costs) of three types of 4-year institutions that I included. This consists of *Public In-State, Public Out-of-state,* and *Private.* Each scatter plot lined up against each other consists of the same variables and numbers for unemployment as well as the tuition costs. The only difference is the type of tuition and the results.

**Results:**

*Private -* The data shown on this chart displays evidence that higher costs of tuition for Private 4-year institutions results to higher rates of unemployment. This is represented as there is a positive line indicating a positive correlation between these factors. The line for the correlations aren't necessarily "strong" as the data plotted is scattered all over and not very dense along the line.

*Public In-State* - The costs of tuition for Public In-State schools are the lowest of the other two types. It also shows a slightly stronger correlation than the other two as well, but the data for this is shown to be a negative correlation with values at an average of about \$10,000.

*Public Out-of-State* - This graphs represents a higher positive correlation for lower unemployment rates with tuition costs, and a slightly lower positive correlation for higher unemployment rates being affect by the tuition cost. I believe this is due to the states original states unemployment rates not being directly affected by higher out of state tuition costs.

**Conclusion:**

As a result of my data, I don't suppose that all types of 4-year institutions have directly affected the unemployment rates throughout the country, but I do recognize a higher positive correlation with this for specifically Private institutions than the others. Due to there not being an extremely strong correlation, I don't believe there is relatively enough data that supports my question. I do believe there are other contributing factors that have stronger impacts and higher correlations on the unemployment levels such as the rise in cost of living, companies relying on AI, and lowering of budgets.
